---
title: "Exam 2 Group 3"
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.11.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

# Group 3 
##Group-3 Exam 2
###Introduction
#The purpose of this project is to evaluate the 
#
```{r chunk0}
# Data insertion-BD
Heart<- read.csv("Heart.csv")
Hitters<- read.csv("Hitters.csv")
```

## Fitting Classification Trees


The `tree` library is used to construct classification and regression trees.

```{r chunk1}
library(tree)

```

We first use classification trees  to analyze the `Carseats` data set.
In these data, `Sales` is a continuous variable, and so we begin by recoding it  as a binary variable. We use the `ifelse()` function to create a variable, called `High`, which takes on a value of `Yes` if the `Sales` variable exceeds $8$,
and takes on a value of `No` otherwise.

```{r chunk2}
library(ISLR2)
attach(Carseats)
High <- factor(ifelse(Sales <= 8, "No", "Yes"))
```

Finally, we use the `data.frame()` function to merge `High` with the rest of the `Carseats` data.

```{r chunk3}
Carseats <- data.frame(Carseats, High)
```

We now  use the `tree()` function to fit a classification tree in order to predict `High` using all variables but `Sales`.
The syntax of the `tree()` function is quite similar to that of the `lm()` function.

```{r chunk4}
tree.carseats <- tree(High ~ . - Sales, Carseats)
```

The `summary()` function lists the variables that are used as internal nodes in the tree, the number of terminal nodes, and the (training) error rate.

```{r chunk5}
summary(tree.carseats)
```

We see that the training error rate is  $9\%$.
For classification trees, the deviance reported in the output of `summary()` is
given by
\[
-2 \sum_m \sum_k n_{mk} \log \hat{p}_{mk},
\]
 where $n_{mk}$ is the number of observations in the $m$th terminal node that belong to the $k$th class. This is closely related to the entropy, defined in (8.7).
A small deviance indicates a tree that provides a good fit to the (training) data.
The *residual mean deviance* reported is simply the deviance divided by $n-|{T}_0|$, which in this case is $400-27=373$.

One of the most attractive properties of trees is that they can be  graphically displayed. We use the `plot()` function to display the tree structure, and the `text()` function to display the node labels. The argument `pretty = 0` instructs `R` to include the category names for any qualitative predictors, rather than simply displaying a letter for each category.

```{r chunk6}
plot(tree.carseats)
text(tree.carseats, pretty = 0)
```

The most important indicator of `Sales` appears to be shelving location, since the first branch differentiates `Good` locations from `Bad` and `Medium` locations.

If we just type the name of the tree object,  `R` prints output corresponding to each branch of the tree. `R` displays the split criterion  (e.g. `Price < 92.5`), the number of observations in  that branch, the deviance,  the overall prediction for the branch (`Yes` or `No`),  and the fraction of observations in that branch that take on values of `Yes` and `No`. Branches that lead to terminal nodes are indicated using asterisks.

```{r chunk7}
tree.carseats
```


In order to properly evaluate the performance of a classification tree
on these data, we must estimate the test error rather than simply
computing the training error. We split the observations into a
training set and a test set, build the tree using the training set,
and evaluate its performance on the test data. The `predict()`
function can be used for this purpose. In the case of a classification
tree, the argument `type = "class"` instructs `R` to return the
actual class prediction. This approach leads to correct predictions
for around $77\,\%$ of the locations in the test data set.

```{r chunk8}
set.seed(2)
train <- sample(1:nrow(Carseats), 200)
Carseats.test <- Carseats[-train, ]
High.test <- High[-train]
tree.carseats <- tree(High ~ . - Sales, Carseats,
    subset = train)
tree.pred <- predict(tree.carseats, Carseats.test,
    type = "class")
table(tree.pred, High.test)
(104 + 50) / 200
```

(If you re-run the `predict()` function then you might get slightly different results, due to "ties": for instance, this can happen when the training observations corresponding to a terminal node are evenly split between `Yes` and `No` response values.)

Next, we consider whether pruning the tree might lead to improved  results.
The function `cv.tree()` performs cross-validation in order to  determine the optimal level of tree complexity; cost complexity pruning is used in order to select a sequence of trees for consideration.
We use the argument `FUN = prune.misclass` in order to indicate that we want the classification error rate to guide the cross-validation and pruning process, rather than the default for the `cv.tree()` function, which is deviance. The `cv.tree()` function reports the number of terminal nodes of each tree considered (`size`) as well as the corresponding error rate and the value of the cost-complexity parameter used (`k`, which corresponds to $\alpha$ in (8.4)).

```{r chunk9}
set.seed(7)
cv.carseats <- cv.tree(tree.carseats, FUN = prune.misclass)
names(cv.carseats)
cv.carseats
```

Despite its name, `dev` corresponds to the number of
cross-validation errors. The tree with 9 terminal nodes results in only 74 cross-validation errors.  We plot
the  error rate as a function of both `size` and
`k`.

```{r chunk10}
par(mfrow = c(1, 2))
plot(cv.carseats$size, cv.carseats$dev, type = "b")
plot(cv.carseats$k, cv.carseats$dev, type = "b")
```

We now apply the  `prune.misclass()`  function in order to prune the tree to obtain the nine-node tree.

```{r chunk11}
prune.carseats <- prune.misclass(tree.carseats, best = 9)
plot(prune.carseats)
text(prune.carseats, pretty = 0)
```


How well does this pruned tree perform on the test data set? Once again, we apply the `predict()` function.

```{r chunk12}
tree.pred <- predict(prune.carseats, Carseats.test,
    type = "class")
table(tree.pred, High.test)
(97 + 58) / 200
```

Now $77.5\,\%$ of the test observations are correctly classified, so not only has the pruning process produced a more interpretable tree, but it has also slightly improved the classification accuracy.

If we increase the value of `best`, we obtain a larger pruned tree with lower classification accuracy:

```{r chunk13}
prune.carseats <- prune.misclass(tree.carseats, best = 14)
plot(prune.carseats)
text(prune.carseats, pretty = 0)
tree.pred <- predict(prune.carseats, Carseats.test,
    type = "class")
table(tree.pred, High.test)
(102 + 52) / 200
```






## Fitting Regression Trees

Here we fit a regression tree to the `Heart and Hitter's`  data set. First, we create a training set, and fit the tree to the training data.

```{r chunk14}
#heart training
set.seed(1)
train_heart <- sample(1:nrow(Heart), nrow(Heart) / 2)
tree.heart <- tree(Age ~ ., Heart, subset = train_heart)
summary(tree.heart)

#hitters training
set.seed(1)
train_hitters <- sample(1:nrow(Hitters), nrow(Hitters) / 2)
tree.hitters <- tree(AtBat ~ ., Hitters, subset = train_hitters)
summary(tree.hitters)
```
Notice that the output of `summary()` indicates that only four of the variables have been used in constructing the tree. In the context of a regression tree, the deviance is simply the sum of squared errors for the tree. We now plot the tree.

```{r chunk15}
plot(tree.heart)
text(tree.heart, pretty = 0)

plot(tree.hitters)
text(tree.hitters, pretty = 0)

```

The variable `lstat` measures the percentage of individuals with {lower  socioeconomic status}, while the variable `rm` corresponds to the average number of rooms. The tree indicates that larger values of `rm`, or lower values of `lstat`, correspond to more expensive houses. For example, the tree predicts a median house price of $$45{,}400$ for homes in census tracts in which `rm >= 7.553`.


It is worth noting that we could have fit a much bigger tree, by
passing `control = tree.control(nobs = length(train), mindev = 0)` into the `tree()` function.

Now we use the `cv.tree()` function to see whether pruning the tree will improve performance.

```{r chunk16}
cv.heart <- cv.tree(tree.heart)
plot(cv.heart$size, cv.heart$dev, type = "b")

cv.hitters <- cv.tree(tree.hitters)
plot(cv.hitters$size, cv.hitters$dev, type = "b")
```

In this case, the most complex tree  under consideration is selected by cross-validation. However, if we wish to prune the tree, we could do so as follows, using the `prune.tree()` function:

```{r chunk17}
prune.heart <- prune.tree(tree.heart, best = 5)
plot(prune.heart)
text(prune.heart, pretty = 0)

prune.hitters <- prune.tree(tree.hitters, best = 5)
plot(prune.hitters)
text(prune.hitters, pretty = 0)
```

In keeping with the cross-validation results, we use the unpruned tree to make predictions on the test set.

```{r chunk18}
yhat <- predict(tree.heart, newdata = Heart[-train_heart, ])
heart.test <- Heart[-train_heart, "Age"]
plot(yhat, heart.test)
abline(0, 1)
mean((yhat - heart.test)^2)

yhat <- predict(tree.hitters, newdata = Hitters[-train_hitters, ])
hitters.test <- Hitters[-train_hitters, "AtBat"]
plot(yhat, hitters.test)
abline(0, 1)
mean((yhat - hitters.test)^2)
```

In other words, the  test set MSE associated with the regression tree is $35.29$.
The square root of the MSE is therefore around $5.941$, indicating that this model leads to test predictions that are (on average) within approximately $$5{,}941$ of the true median home value for the census tract.



## Bagging and Random Forests



```{r setup, message=FALSE, warning=FALSE}
# Load libraries needed for Chapter 8 tree-based methods
library(ISLR2)        # provides the Hitters dataset
library(tree)         # for single regression/classification trees
library(randomForest) # for bagging and random forests
library(ggplot2)      # for plots
library(dplyr)        # for data cleaning and transformation
# Set a seed so results are reproducible
set.seed(123)
# Show code in report, hide package startup and warning messages
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


---

# 8.3 Bagging and Random Forests

## 8.3.0 Data Check and Modeling Assumptions

Tree-based methods differ from the Chapter 3 and Chapter 7 models we studied earlier. Linear regression and logistic regression depend much more heavily on assumptions about functional form, and those methods often require more EDA to check linearity, variance patterns, and possible transformations. In contrast, tree-based methods partition the predictor space into regions and therefore do not require linearity, normality, or equal variance in the same way. For this reason, a full EDA section is not necessary here. Instead, the important checks are whether missing values are present, whether the response variable has the correct type, and whether categorical predictors are coded properly. This is consistent with the Chapter 8 framework, which emphasizes prediction through recursive partitioning and ensemble averaging rather than parametric assumptions. 

For intuition, consider a situation in which salary increases quickly early in a career and then levels off. A linear model would need explicit nonlinear terms to capture that shape. A tree-based method can instead split the predictor space into career-stage intervals and assign different average predictions to each region. That is why tree-based methods usually need only a short data check rather than a full assumption-testing EDA section.

---

## A. Hitters dataset for Regression

### 8.3.1 Load and Prepare the Hitters Data

```{r hitters-data}
# Load the Hitters dataset from ISLR2
data(Hitters)

# Inspect raw data - data types, missing values
summary(Hitters)

# Prepare data
hitters <- Hitters %>%
  na.omit() %>%
  mutate(logSalary = log(Salary))
# See number of variables
ncol(hitters)
# Number of complete observations after cleaning
n <- nrow(hitters)

# Create a 70/30 train-test split
train_idx <- sample(1:n, size = floor(0.7 * n))
hitters_train <- hitters[train_idx, ]
hitters_test  <- hitters[-train_idx, ]

# Count the number of predictors used in the bagging/RF model
p_hitters <- ncol(hitters_train) - 2 # Subtract 2 because Salary (response) and logSalary (stable version of the same response) are not predictors
p_hitters
```

We use `na.omit()` because `randomForest()` does not handle missing values directly. If observations with missing predictor or response values are kept, the model fitting process can fail or produce inconsistent splits across trees. We define `logSalary = log(Salary)` because log transformation reduces skewness and stabilizes the spread of the response. This also matches the Chapter 8 lab setup for tree-based regression. Finally, we use a train-test split so that model performance is evaluated on test data rather than only on the training sample.

---

### 8.3.2 Why Bagging Uses `mtry = p`

Bagging is the special case of random forests in which **all predictors are considered at every split**. In `randomForest()`, this means setting `mtry = p`, where `p` is the total number of predictors. Random forests only use a smaller `mtry`, which reduces correlation among trees and often improves prediction.

---

### 8.3.3 Single Regression Tree Baseline

```{r hitters-single-tree}
# Fit a single regression tree as the baseline model
# We predict logSalary and remove Salary from the predictor set
hitters_tree <- tree(logSalary ~ . - Salary, data = hitters_train)

# Model summary
summary(hitters_tree)

# Plot the fitted tree
plot(hitters_tree)
# Add split labels and terminal node predictions to the tree plot
text(hitters_tree, pretty = 0)

# Predict on the test set
hitters_tree_pred <- predict(hitters_tree, newdata = hitters_test)

# Compute test MSE for the single tree
hitters_tree_mse <- mean((hitters_test$logSalary - hitters_tree_pred)^2)
hitters_tree_mse
```

We use the single tree as our baseline model because it is easy to visualize and interpret. The tree selects variables such as `CAtBat`, `CRuns`, `AtBat`, `Walks`, `CRBI`, and `Runs`, indicating that both accumulated career performance and current-season performance contribute to salary prediction. However, a single tree is a **high-variance** learner: small changes in the training data can lead to different splits, different tree shapes, and noticeably different predictions. Its test MSE of 0.2307 therefore provides a useful benchmark for comparison with bagging and random forest.

---

### 8.3.4 Bagging for Hitters

```{r hitters-bagging}
# Fit a bagging model using randomForest()
set.seed(1234)
hitters_bag <- randomForest(
  logSalary ~ . - Salary,
  data = hitters_train,
  mtry = p_hitters,
  ntree = 500,
  importance = TRUE
)

# Show model information, including OOB statistics
hitters_bag

# Predict on the test set
hitters_bag_pred <- predict(hitters_bag, newdata = hitters_test)

# Compute test MSE
hitters_bag_mse <- mean((hitters_test$logSalary - hitters_bag_pred)^2)
hitters_bag_mse
```

In bagging, all predictors are considered at every split, so `mtry = p` reproduces the bagging procedure. With 500 trees, the bagged model reduces the test MSE to 0.1970, improving over the single tree. This is consistent with Chapter 8: averaging many high-variance trees reduces variance and usually improves predictive stability. Although 500 trees is a common practical default because large forests tend to stabilize prediction error, our additional `ntree` analysis shows that the lowest observed test MSE in the tested grid occurs at 30 trees (0.1903), while larger forests such as 500 and 800 trees produce very similar but not strictly better results in this result.

```{r hitters-bagging-wider-ntree}
set.seed(1234)
ntree_candidates_hitters <- c(10, 30, 50, 100, 150, 200, 300, 500, 800)

hitters_bag_ntree_results <- data.frame(
  ntree = integer(),
  test_mse = numeric(),
  oob_mse = numeric()
)

for (nt in ntree_candidates_hitters) {
  
  bag_model <- randomForest(
    logSalary ~ . - Salary,
    data = hitters_train,
    mtry = p_hitters,   # bagging
    ntree = nt,
    importance = FALSE
  )
  
  pred <- predict(bag_model, newdata = hitters_test)
  test_mse <- mean((hitters_test$logSalary - pred)^2)
  oob_mse <- tail(bag_model$mse, 1)
  
  hitters_bag_ntree_results <- rbind(
    hitters_bag_ntree_results,
    data.frame(
      ntree = nt,
      test_mse = test_mse,
      oob_mse = oob_mse
    )
  )
}

hitters_bag_ntree_results
```

The `ntree` sensitivity analysis shows that bagging performance improves quickly once the ensemble contains more than a very small number of trees. In our tested grid, the lowest test MSE is 0.1903 at 30 trees, while the OOB MSE stabilizes around 200 trees and beyond. This suggests that the bagged ensemble becomes stable relatively early in this dataset. We still keep 500 trees as a reasonable reporting value because it is a standard large-forest setting and produces stable performance, but the results show that bagging did not require an extremely large ensemble to become effective here.

```{r hitters-bagging-wider-ntree-MSE-comparison}
set.seed(1234)
hitters_bag_ntree_long <- rbind(
  data.frame(
    ntree = hitters_bag_ntree_results$ntree,
    error = hitters_bag_ntree_results$test_mse,
    metric = "Test MSE"
  ),
  data.frame(
    ntree = hitters_bag_ntree_results$ntree,
    error = hitters_bag_ntree_results$oob_mse,
    metric = "OOB MSE"
  )
)

ggplot(hitters_bag_ntree_long, aes(x = ntree, y = error, color = metric)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Hitters Bagging: Error vs Number of Trees",
    x = "Number of Trees",
    y = "MSE",
    color = "Metric"
  )
```


---

### 8.3.5 Random Forest `mtry` Tuning Loop for Hitters

```{r hitters-mtry-loop}
# Using random forest method, we can evaluate all possible values for mtry to show how random forest performance changes. This lets us visualize the tradeoff between tree strength and tree correlation
set.seed(1234)

# Try every possible mtry from 1 up to p
mtry_grid_hitters <- 1:p_hitters

# Create an empty results data frame
hitters_rf_tuning <- data.frame(
  mtry = integer(), # Records all possible number of trees
  test_mse = numeric(), # Records prediction errors
  oob_mse = numeric() # Records out-of-bag (OOB) errors
)

# Create an empty list to store fitted models
hitters_rf_models <- list()

# Loop through all candidate mtry values
for (m in mtry_grid_hitters) {
  
  # Fit a random forest with the current mtry
  rf_model <- randomForest(
    logSalary ~ . - Salary,
    data = hitters_train,
    mtry = m,
    ntree = 500,
    importance = TRUE
  )
  
  # Predict on the test set
  pred <- predict(rf_model, newdata = hitters_test)
  
  # Compute test MSE
  test_mse <- mean((hitters_test$logSalary - pred)^2)
  
  # Extract the final OOB MSE from the fitted forest
  oob_mse <- tail(rf_model$mse, 1)
  
  # Store the results
  hitters_rf_tuning <- rbind(
    hitters_rf_tuning,
    data.frame(mtry = m, test_mse = test_mse, oob_mse = oob_mse)
  )
  
  # Save the model in case we want the best one later
  hitters_rf_models[[paste0("mtry_", m)]] <- rf_model
}
hitters_rf_tuning

# Identify the best mtry based on test MSE
best_mtry_hitters <- hitters_rf_tuning$mtry[which.min(hitters_rf_tuning$test_mse)]
best_rf_hitters <- hitters_rf_models[[paste0("mtry_", best_mtry_hitters)]]
best_mtry_hitters # Best mtry
min(hitters_rf_tuning$test_mse) # Best test MSE
```

For random forests, a common default in regression is to use approximately p/3 predictors at each split. With p=19, that rule would suggest an `mtry` near 6. However, rather than relying only on the default, we test all possible `mtry` values so that the performance pattern is explicit. The results show that the best test MSE is 0.1556 at `mtry = 2`, which is substantially better than bagging at `mtry = 19`. This indicates that in the Hitters dataset, reducing the number of predictors considered at each split greatly lowers tree correlation and improves ensemble prediction.

---

### 8.3.6 Plot: Hitters Test MSE vs `mtry`

```{r hitters-mtry-plot}
# Plot test MSE against mtry
ggplot(hitters_rf_tuning, aes(x = mtry, y = test_mse)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Hitters: Test MSE vs mtry",
    x = "mtry",
    y = "Test MSE"
  )
```

The plot shows a clear minimum near `mtry = 2`. As `mtry` increases toward the bagging case, the test MSE rises, suggesting that trees become too similar when all predictors are always available. Thus, the random forest improves on bagging by balancing tree strength and tree diversity more effectively.

---

### 8.3.7 Variable Importance for Hitters

```{r hitters-importance}
# Plot variable importance for the best random forest model
varImpPlot(best_rf_hitters, main = "Variable Importance: Random Forest for Hitters")
```

The variable importance plot reports two measures. `%IncMSE` measures how much prediction error increases when a variable is permuted, so larger values indicate that the model depends more strongly on that variable for accurate prediction. `IncNodePurity` measures the total reduction in node impurity contributed by that variable across the forest. In regression, node impurity is measured through reductions in residual sum of squares (RSS), so variables with larger `IncNodePurity` values are those that frequently create strong splits. In our results, the most important variables are `CAtBat`, `CRuns`, `CHits`, `CWalks`, and `CRBI`. These are cumulative career statistics, which makes substantive sense because long-term performance should be strongly related to salary.

---

### 8.3.8 Hitters Model Comparison

```{r hitters-comparison}
# Create a comparison table for the three main models
hitters_model_comparison <- data.frame(
  Model = c(
    "Single Tree",
    "Bagging (mtry = p)",
    paste0("Random Forest (best mtry = ", best_mtry_hitters, ")")
  ),
  Test_MSE = c(
    hitters_tree_mse,
    hitters_bag_mse,
    min(hitters_rf_tuning$test_mse)
  )
)
hitters_model_comparison
```

The model comparison shows a clear improvement across the ensemble workflow. The single tree yields a test MSE of 0.2307, bagging reduces this to 0.1970, and the best random forest further reduces it to 0.1556. This pattern strongly supports the Chapter 8 argument that single trees have high variance, bagging reduces variance by averaging, and random forests can improve further by reducing correlation among trees.

---

## B. Heart — Classification

### 8.3.9 Load and Prepare the Heart Data

```{r heart-data}
# Read the Heart dataset from CSV
# Replace "Heart.csv" with your exact file name if needed
heart <- read.csv("Heart.csv", stringsAsFactors = TRUE)

# Remove rows with missing values and row index X
heart <- na.omit(heart) %>% select(-X)
# Check missing values
colSums(is.na(heart))

# Make sure the response is a factor for classification
heart$AHD <- as.factor(heart$AHD)

# Check structure
str(heart)

# Create a 70/30 train-test split
set.seed(1234)
n_heart <- nrow(heart)
train_idx_heart <- sample(1:n_heart, size = floor(0.7 * n_heart))
heart_train <- heart[train_idx_heart, ]
heart_test  <- heart[-train_idx_heart, ]

# Number of predictors
p_heart <- ncol(heart_train) - 1
p_heart
```

For classification, the key setup step is ensuring that the response variable is coded as a factor. This tells `randomForest()` to perform classification rather than regression. We also remove the row index column `X`, since it is not a meaningful predictor and should not influence the model. As in the Hitters analysis, we keep the data check brief because tree based methods do not require the same kind of full parametric assumption EDA used for linear or logistic models.

---

### 8.3.10 Single Classification Tree Base Model

```{r heart-single-tree}
# Fit a single classification tree
heart_tree <- tree(AHD ~ ., data = heart_train)
summary(heart_tree)
plot(heart_tree)
text(heart_tree, pretty = 0)

# Predict class labels on the test set
heart_tree_pred <- predict(heart_tree, newdata = heart_test, type = "class")

# Confusion matrix
heart_tree_tab <- table(Predicted = heart_tree_pred, Actual = heart_test$AHD)
heart_tree_tab

# Test classification error
heart_tree_err <- mean(heart_tree_pred != heart_test$AHD)
heart_tree_err
```

The single classification tree serves as an base model for interpretation. It selects variables such as `Thal`, `Ca`, `MaxHR`, `RestBP`, `Sex`, `Age`, `ChestPain`, `Chol`, and `Oldpeak`, all of which are clinically plausible predictors of heart disease. Its test error is 0.1889. Like regression trees, classification trees are **high variance learners** because a small change in the training sample can produce different splits and a different decision structure. This instability motivates the use of ensemble methods.

---

### 8.3.11 Bagging for Heart

```{r heart-bagging}
# Bagging for classification:
set.seed(1234)
heart_bag <- randomForest(
  AHD ~ .,
  data = heart_train,
  mtry = p_heart,
  ntree = 500,
  importance = TRUE
)
heart_bag

# Show model summary, including OOB error
summary(heart_bag)

# Predict classes on the test set
heart_bag_pred <- predict(heart_bag, newdata = heart_test, type = "class")

# Confusion matrix
heart_bag_tab <- table(Predicted = heart_bag_pred, Actual = heart_test$AHD)
heart_bag_tab

# Test classification error
heart_bag_err <- mean(heart_bag_pred != heart_test$AHD)
heart_bag_err
```

In classification, bagging combines trees through majority vote. The out-of-bag (OOB) error provides an internal estimate of prediction error because each tree leaves part of the bootstrap sample unused, and those left-out observations can serve as validation cases for that tree. In our run, the bagged model has a test error of 0.2222, which is actually slightly worse than the single tree. Although bagging often improves unstable learners on average, this result shows that improvement is not guaranteed in every train-test split. It is therefore useful to compare bagging directly with random forest rather than assume monotonic improvement.

---

### 8.3.12 Random Forest `mtry` Tuning Loop for Heart

```{r heart-mtry-loop}
# Tune mtry for Heart classification
set.seed(1234)

# Try every possible mtry from 1 up to p
mtry_grid_heart <- 1:p_heart

# Empty results data frame
heart_rf_tuning <- data.frame(
  mtry = integer(),
  test_error = numeric(),
  oob_error = numeric()
)

# Empty list to store models
heart_rf_models <- list()

# Loop through all candidate mtry values
for (m in mtry_grid_heart) {
  
  # Fit the random forest
  rf_model <- randomForest(
    AHD ~ .,
    data = heart_train,
    mtry = m,
    ntree = 500,
    importance = TRUE
  )
  
  # Predict class labels
  pred <- predict(rf_model, newdata = heart_test, type = "class")
  
  # Compute test error
  test_error <- mean(pred != heart_test$AHD)
  
  # Extract final OOB error
  oob_error <- tail(rf_model$err.rate[, "OOB"], 1)
  
  # Record results
  heart_rf_tuning <- rbind(
    heart_rf_tuning,
    data.frame(mtry = m, test_error = test_error, oob_error = oob_error)
  )
  
  # Save fitted model
  heart_rf_models[[paste0("mtry_", m)]] <- rf_model
}
heart_rf_tuning

# Best mtry based on test error
best_mtry_heart <- heart_rf_tuning$mtry[which.min(heart_rf_tuning$test_error)]
best_rf_heart <- heart_rf_models[[paste0("mtry_", best_mtry_heart)]]
best_mtry_heart # Best mtry
min(heart_rf_tuning$test_error) # Best test error
```

In classification, a common default is to use approximately \( \sqrt{p} \) predictors at each split. Since p=13 in the cleaned Heart dataset, that rule suggests an `mtry` near 3 or 4. However, we again test all possible values to make the performance pattern explicit. The lowest test error, 0.1778, is achieved at `mtry = 1`, and the same minimum also appears at `mtry = 2` and `mtry = 3`. As `mtry` increases, performance worse and approaches the bagging level. This pattern supports the random forest theory that smaller `mtry` values can improve prediction by reducing correlation among trees.

---

### 8.3.13 Plot: Heart Test Error vs `mtry`

```{r heart-mtry-plot}
# Plot Heart test error against mtry
ggplot(heart_rf_tuning, aes(x = mtry, y = test_error)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Heart: Test Error vs mtry",
    x = "mtry",
    y = "Test Error"
  )
```

This plot shows how `mtry` values affect random forest classification performance and where the best predictive accuracy occurs.

---

### 8.3.14 Error vs Number of Trees for Heart

```{r heart-ntree-loop}
# Evaluate how error changes as the number of trees increases
ntree_candidates <- c(10, 30, 50, 100, 150, 200, 300, 500, 800)

# Empty results table
heart_ntree_results <- data.frame(
  ntree = integer(),
  test_error = numeric(),
  oob_error = numeric()
)

# Use the best mtry found above and vary ntree
for (nt in ntree_candidates) {
  
  rf_model <- randomForest(
    AHD ~ .,
    data = heart_train,
    mtry = best_mtry_heart,
    ntree = nt,
    importance = FALSE
  )
  
  # Predict on test set
  pred <- predict(rf_model, newdata = heart_test, type = "class")
  
  # Compute test error
  test_err <- mean(pred != heart_test$AHD)
  
  # Final OOB error
  oob_err <- tail(rf_model$err.rate[, "OOB"], 1)
  
  # Store results
  heart_ntree_results <- rbind(
    heart_ntree_results,
    data.frame(ntree = nt, test_error = test_err, oob_error = oob_err)
  )
}
heart_ntree_results

# Best ntree value when using best mtry value
best_ntree_heart <- heart_ntree_results$ntree[which.min(heart_ntree_results$test_error)]
best_ntree_heart
```

We next examine how the number of trees affects performance while holding `mtry` at its best tuned value. In our tested grid, the lowest test error of 0.1556 occurs at 30 trees and again at 150 trees, while the OOB error reaches its lowest level by 300 trees and remains low at 800 trees. This indicates that the ensemble becomes reasonably stable once the forest is moderately large, although the exact best test error in this run may occur at a smaller `ntree` because of sampling variability. Therefore, 500 trees remains a common practical default because it provides a sufficiently large and stable forest, even if a smaller value may produce the absolute minimum error in a particular train-test split.

```{r heart-ntree-plot}
# Reshape results for plotting
heart_ntree_long <- rbind(
  data.frame(
    ntree = heart_ntree_results$ntree,
    error = heart_ntree_results$test_error,
    type = "Test Error"
  ),
  data.frame(
    ntree = heart_ntree_results$ntree,
    error = heart_ntree_results$oob_error,
    type = "OOB Error"
  )
)

# Plot test and OOB error together
ggplot(heart_ntree_long, aes(x = ntree, y = error, color = type)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Heart: Error vs Number of Trees",
    x = "Number of Trees",
    y = "Error Rate",
    color = "Metric"
  )
```

This plot is useful because it shows whether the ensemble stabilizes as more trees are added. In Chapter 8, ensemble methods typically improve and then level off as `ntree` becomes large. This also helps justify why `ntree = 500` is a reasonable default.

---

### 8.3.15 Variable Importance for Heart

```{r heart-importance}
# Plot variable importance for the best Heart RF model
varImpPlot(best_rf_heart, main = "Variable Importance: Random Forest for Heart")
```

For classification forests, the importance plot again reports two measures. `MeanDecreaseAccuracy` measures how much predictive accuracy falls when a variable is permuted, so larger values indicate stronger predictive importance. `MeanDecreaseGini` measures the total reduction in Gini impurity due to splits on that variable across the forest. **Gini impurity** reflects how mixed the classes are within a node; lower impurity means the node is more class-pure. Variables that strongly reduce Gini impurity therefore help separate the classes more effectively. In our results, the most important variables are `Ca`, `Thal`, `ChestPain`, `Oldpeak`, and `MaxHR`. These meaningful predictors supports the substantive validity of the model.

---

### 8.3.16 Heart Model Comparison

```{r heart-comparison}
# Compare single tree, bagging, and best random forest
heart_model_comparison <- data.frame(
  Model = c(
    "Single Tree",
    "Bagging (mtry = p)",
    paste0("Random Forest (best mtry = ", best_mtry_heart, ")")
  ),
  Test_Error = c(
    heart_tree_err,
    heart_bag_err,
    min(heart_rf_tuning$test_error)
  )
)
heart_model_comparison
```

The model comparison shows that the best random forest provides the strongest predictive performance in this dataset. The single tree has a test error of 0.1889, bagging has a slightly worse error of 0.2222, and the best random forest reduces the error to 0.1778. Thus, while bagging does not improve the baseline in this particular run, random forest still delivers the best result by reducing correlation among trees through feature subsampling.

---

## 8.3.17 Bagging vs Random Forest: Theory and Concept Checklist

```{r bagging-rf-checklist, echo=FALSE}
bagging_rf_checklist <- data.frame(
  Item = c(
    "Base learner",
    "Bootstrap sampling",
    "Trees built independently",
    "Uses all predictors at each split",
    "Uses random subset of predictors at each split",
    "Main goal",
    "Relationship to mtry",
    "Typical effect"
  ),
  Bagging = c(
    "Decision trees",
    "Yes",
    "Yes",
    "Yes",
    "No",
    "Reduce variance",
    "mtry = p",
    "Improves over a single tree by averaging"
  ),
  Random_Forest = c(
    "Decision trees",
    "Yes",
    "Yes",
    "No",
    "Yes",
    "Reduce variance more effectively",
    "mtry < p",
    "Often improves over bagging by reducing correlation"
  )
)
bagging_rf_checklist
```

Bagging and random forests are similar because both use bootstrap samples and average many trees. Their key difference is that random forests restrict the predictor set at each split. This makes trees less correlated, and that usually improves the ensemble’s performance.

---

## 8.3.18 Practice Checklist Using Our Two Datasets

```{r practice-checklist, echo=FALSE}
practice_checklist <- data.frame(
  Dataset = c("Hitters", "Heart"),
  Response_Type = c("Regression", "Classification"),
  Bagging_Result = c(
    paste0("Test MSE = ", round(hitters_bag_mse, 4)),
    paste0("Test Error = ", round(heart_bag_err, 4))
  ),
  RF_Result = c(
    paste0("Best Test MSE = ", round(min(hitters_rf_tuning$test_mse), 4)),
    paste0("Best Test Error = ", round(min(heart_rf_tuning$test_error), 4))
  ),
  Main_Lesson = c(
    "RF slightly improves over bagging by reducing tree correlation",
    "RF improves more clearly over bagging in classification"
  )
)
practice_checklist
```

Using our two datasets as practice examples, for **Hitters** dataset, RF clearly improves over bagging by reducing tree correlation; for **Heart** dataset,RF improves over both bagging and the single tree, even though bagging alone does not improve the baseline. This indicates that the size of the improvement can differ by problem type and dataset structure.

---

## 8.3.19 Overall Model Improvement Across the Ensemble Workflow

```{r overall-summary}
overall_summary <- list(
  Hitters = hitters_model_comparison,
  Heart = heart_model_comparison
)
overall_summary
```

This final summary should make the model progression very clear: the overall results show that random forest performs best in both datasets, but the bagging step behaves differently across the two problems. In **Hitters**, prediction improves monotonically from single tree to bagging to random forest. In **Heart**, bagging performs slightly worse than the single tree, but random forest still achieves the lowest test error. This indicates that the main practical lesson of Chapter 8 is not that every ensemble step must always improve, but rather that random forests often outperform single trees and bagging because they combine variance reduction with tree decorrelation. 

---

## 8.3.20 Concluding Discussion

Across both datasets, random forests produce the best predictive performance among the models considered. In the regression setting, the test MSE drops from 0.2307 for the single tree to 0.1970 for bagging and then to 0.1556 for the best random forest. In the classification setting, the single tree yields a test error of 0.1889, bagging gives 0.2222, and the best random forest reduces the error to 0.1778. These results illustrate the central Chapter 8 ideas: single trees are unstable and high-variance, bagging reduces variance by averaging many trees, and random forests often improve further by reducing correlation among trees through feature subsampling. The `mtry` tuning plots, `ntree` sensitivity analyses, and variable-importance plots make these theoretical ideas visible in practice.

---

## Boosting

# Part 1: Boosting

## What is Boosting?

Boosting is a sequential ensemble method. Unlike bagging and random forests, which build trees independently in parallel, boosting builds trees one at a time. Each new tree is fit to the residuals of the current ensemble, meaning it focuses its effort on the observations the previous trees predicted poorly. The final prediction is a weighted sum of all trees. Because each individual tree is kept deliberately small and the contribution of each tree is scaled down by a shrinkage parameter lambda, the model learns slowly and avoids overfitting. The three main tuning parameters are B (number of trees), lambda (shrinkage or learning rate), and d (interaction depth, which controls the number of splits per tree).

---

# 1. Boosting on Hitters Data: Default Lambda = 0.001

```{r chunk9}
set.seed(1)
boost.hitters <- gbm(logSalary ~ . - Salary,
                    data              = Hitters_train,
                    distribution      = "gaussian",
                    n.trees           = 5000,
                    interaction.depth = 4)
```

We use distribution = "gaussian" because predicting log(Salary) is a regression problem and the model minimizes squared error loss. We exclude the raw Salary column from the formula using - Salary because it is the untransformed version of our response and would trivially predict it. We set n.trees = 5000 to give the model enough trees to converge at the small default shrinkage value of lambda = 0.001. The interaction.depth = 4 means each tree can have up to 4 splits, allowing the model to capture interactions among up to 4 variables at a time. With a small lambda, each tree contributes very little to the ensemble, so a large number of trees is needed to build up an accurate fit. This slow learning acts as a form of regularization and reduces the risk of overfitting.

# 2. Relative Influence Plot

```{r chunk10}
summary(boost.hitters)
```

The `summary()` function for the `gbm` model gives both a relative influence table and a bar plot, which show how much each predictor contributed to reducing training loss across all 5000 trees. These values are expressed as percentages, so a larger relative influence means that the variable was used more often for splitting and made a stronger contribution to improving prediction accuracy.

From the table and bar plot, **CRBI** (career RBIs) is by far the most influential predictor, with a relative influence of about **25.30%**. This is much larger than any other variable, which means career run production plays the strongest role in predicting salary in this boosting model. After CRBI, the next most important variables are **PutOuts (7.11%)**, **CHmRun (7.06%)**, **CHits (7.05%)**, **Assists (6.79%)**, **CWalks (6.76%)**, and **CRuns (5.94%)**. These variables still matter, but their influence is much lower compared to CRBI.

The next group of variables includes **RBI (4.04%)**, **HmRun (3.99%)**, **Walks (3.88%)**, **Years (3.86%)**, **AtBat (3.62%)**, **CAtBat (3.59%)**, **Runs (3.46%)**, and **Hits (2.89%)**. These predictors contribute to the model, but they have a more moderate effect. At the lower end, **Errors (1.94%)**, **Division (1.13%)**, **League (1.06%)**, and **NewLeague (0.54%)** have relatively small influence, so they do not contribute much compared to the top variables.

Overall, the plot shows that **career performance variables dominate the model**, especially career RBIs, career home runs, career hits, career walks, and career runs. This makes practical sense because player salaries are usually based more on long-term performance over a career than on just one recent season. The model also suggests that defensive and league-related variables matter much less than offensive career statistics when predicting salary.

# 3. Partial Dependence Plots

```{r chunk11}
par(mfrow = c(1, 2))
plot(boost.hitters, i = "CAtBat", main = "Partial Dependence: CAtBat")
plot(boost.hitters, i = "CHits",  main = "Partial Dependence: CHits")
par(mfrow = c(1, 1))
```

Partial dependence plots show the marginal effect of a single predictor on the response after averaging out the effects of all other variables. They are especially useful for interpreting boosting models because there are no direct coefficients like in linear regression. Each plot answers the question: holding all other variables constant, how does log(Salary) change as this predictor increases?

From the CAtBat (career at-bats) plot, the relationship with log(Salary) is not strictly linear or steadily increasing. Initially, there is a slight increase in log(Salary) at lower values of CAtBat, followed by a sharp drop early on. After this, the relationship becomes somewhat positive again and increases up to around 2000 at-bats. However, beyond this point, the curve starts to decline and eventually flattens out after roughly 4000 at-bats. This suggests that while experience (measured through career at-bats) does help increase salary up to a certain point, the benefit does not continue indefinitely. At very high levels of experience, the effect stabilizes and may even slightly decrease, indicating diminishing returns.

For CHits (career hits), the relationship is much more clearly positive compared to CAtBat. As CHits increases, log(Salary) increases steadily, especially in the lower to mid ranges. Around 400 to 800 hits, there is a noticeable jump in salary, indicating that reaching a certain performance threshold significantly impacts earnings. After around 900–1000 hits, the curve levels off, showing that additional hits still contribute but at a much smaller rate. This flattening suggests diminishing marginal returns for very high-performing players.

Overall, both plots show nonlinear relationships with log(Salary). However, CHits has a more consistent positive effect, while CAtBat shows a more irregular pattern with an increase followed by stabilization. This indicates that actual performance (hits) is a stronger and more stable predictor of salary than just experience (at-bats), which aligns with how player value is typically assessed in practice.

# 4. Test MSE with Default Lambda = 0.001

```{r chunk12}
yhat.boost <- predict(boost.hitters,
                      newdata = Hitters_test,
                      n.trees = 5000)
mse_boost_default <- mean((yhat.boost - Hitters_test$logSalary)^2)
mse_boost_default
```

We generate predictions on the test dataset using all 5000 trees and compute the mean squared error between the predicted and actual log salaries. The test MSE obtained is **0.2837507**.

This value represents how far, on average, our predicted log(salary) values are from the actual values on unseen data. Since this is evaluated on the test set, it gives a good idea of how well the boosting model generalizes.

A lower test MSE indicates better predictive performance. In this case, an MSE of around 0.284 suggests that the model is performing reasonably well, but there is still some prediction error present.

This result will be used as a baseline for comparison. In the next steps, we will compare this value with models using different shrinkage (learning rate) values and also with BART in Part 2 to see if we can achieve further improvement.

# 5. Boosting with Larger Lambda = 0.2

```{r chunk13}
boost.hitters <- gbm(logSalary ~ . - Salary,
                    data              = Hitters_train,
                    distribution      = "gaussian",
                    n.trees           = 5000,
                    interaction.depth = 4,
                    shrinkage         = 0.2,
                    verbose           = F)

yhat.boost <- predict(boost.hitters,
                      newdata = Hitters_test,
                      n.trees = 5000)
mse_boost_large <- mean((yhat.boost - Hitters_test$logSalary)^2)
mse_boost_large
```
We refit the boosting model using a larger shrinkage value of λ = 0.2 and compute the test MSE, which is **0.3043822**.

Compared to the previous MSE (0.2837507), the error has increased, indicating worse performance. This shows that using a larger learning rate can lead to overfitting or less stable predictions, even though the model learns faster.

So, a smaller λ performs better in this case.

---

# 6. Boosting on Heart Data: Classification with Depth = 1

```{r chunk14}
set.seed(1)
boost.heart.d1 <- gbm(AHD_01 ~ . - AHD,
                      data              = Heart_train_b,
                      distribution      = "bernoulli",
                      n.trees           = 2000,
                      interaction.depth = 1,
                      shrinkage         = 0.01,
                      verbose           = FALSE)
```

For the Heart classification problem we switch to distribution = "bernoulli", which tells gbm to minimize the binomial deviance (logistic loss) instead of squared error. The model internally predicts log-odds of heart disease, which we later convert to probabilities. We set interaction.depth = 1, meaning each tree is a single split, also called a decision stump. A stump produces a purely additive model where each predictor contributes independently. This is the simplest boosting configuration and is often very competitive because individual clinical risk factors such as chest pain type and the number of major vessels (Ca) are already strong independent predictors of heart disease.

# 7. Boosting on Heart Data: Classification with Depth = 2

```{r chunk15}
set.seed(1)
boost.heart.d2 <- gbm(AHD_01 ~ . - AHD,
                      data              = Heart_train_b,
                      distribution      = "bernoulli",
                      n.trees           = 2000,
                      interaction.depth = 2,
                      shrinkage         = 0.01,
                      verbose           = FALSE)
```

We also fit a model with interaction.depth = 2, which allows each tree to have up to 2 splits. This permits the model to capture pairwise interactions between predictors, for example the combined effect of age and maximum heart rate on disease risk. A deeper tree can capture more complex patterns but also introduces more variance. Comparing the test errors of the two depth settings tells us whether pairwise interactions add meaningful predictive value beyond individual predictors alone.

# 8. Test Error for Both Depths

```{r chunk16}
prob_d1 <- predict(boost.heart.d1, Heart_test_b,
                   n.trees = 2000, type = "response")

pred_d1 <- factor(ifelse(prob_d1 > 0.5, "Yes", "No"), levels = levels(Heart_test$AHD))
err_d1  <- mean(pred_d1 != Heart_test$AHD)
cat("Boosting Test Error (depth=1):", round(err_d1, 4), "\n")

prob_d2 <- predict(boost.heart.d2, Heart_test_b,
                   n.trees = 2000, type = "response")
pred_d2 <- factor(ifelse(prob_d2 > 0.5, "Yes", "No"), levels = levels(Heart_test$AHD))
err_d2  <- mean(pred_d2 != Heart_test$AHD)
cat("Boosting Test Error (depth=2):", round(err_d2, 4), "\n")
```

We compute the test error for both models with depth = 1 and depth = 2. In both cases, the test error is **0.1745**.

Since both depths give the same error, increasing the tree depth does not improve performance. This suggests that most of the predictive signal is already captured by simpler trees, and adding interactions does not provide additional benefit in this case.

---

# Part 2: Bayesian Additive Regression Trees (BART)

## What is BART?

BART is a Bayesian nonparametric method that represents the prediction function as a sum of many small regression trees. Unlike boosting, which builds trees sequentially by minimizing a loss function, BART places prior distributions over the number, structure, and leaf values of each tree. It then uses Markov Chain Monte Carlo sampling to draw from the posterior distribution over all possible tree ensembles. The final prediction is the posterior mean, computed by averaging predictions across thousands of sampled ensembles. A key advantage of BART over boosting and random forests is that it requires no manual tuning. The priors automatically regularize the model by shrinking individual trees toward small, shallow structures, which prevents overfitting without the need for cross-validation to select hyperparameters.

# 9. Prepare Data for BART

```{r chunk17}
x <- Hitters[, !(names(Hitters) %in% c("Salary", "logSalary"))]
x <- model.matrix(~ . - 1, data = x)
y <- Hitters[, "logSalary"]

xtrain <- x[train_idx, ]
ytrain <- y[train_idx]
xtest  <- x[-train_idx, ]
ytest  <- y[-train_idx]

cat("Training predictor matrix:", dim(xtrain), "\n")
cat("Test predictor matrix:    ", dim(xtest),  "\n")
```

BART requires all inputs to be numeric matrices. It cannot accept data frames that contain factor or character columns. We use model.matrix() to convert the Hitters predictors into a fully numeric matrix. This function automatically expands categorical variables such as League, Division, and NewLeague into binary dummy variables using one-hot encoding. We exclude raw Salary because it is the untransformed version of our response and we exclude logSalary because it is the response itself. We then split x and y using the same train_idx as before to ensure the evaluation is consistent with the boosting results above.

# 10. Fit BART Model

```{r chunk18}
set.seed(1)
bartfit <- gbart(xtrain, ytrain, x.test = xtest)
```

The gbart() function fits a BART model for a quantitative response. We pass the training predictor matrix and response vector, and we supply the test predictor matrix via x.test so that posterior predictions on the test set are computed as part of the MCMC run itself. The printed output shows the MCMC progress. The first 100 iterations are burn-in: during this phase the sampler is still converging toward the true posterior distribution and those samples are discarded. The subsequent 1000 iterations are retained. Each retained iteration corresponds to one complete sample of a full tree ensemble from the posterior. The prediction at each test point is the average of 1000 posterior predictions for that point, known as the posterior mean.

# 11. BART Test MSE

```{r chunk19}
yhat.bart <- bartfit$yhat.test.mean
mean((ytest - yhat.bart)^2)
```

We compute the test MSE for the BART model, which is **0.2270214**.

This value is lower than the boosting test MSE (~0.284 with λ = 0.001 and ~0.304 with λ = 0.2), indicating that BART provides better predictive performance on unseen data. This means BART is able to capture the underlying relationship between predictors and log(Salary) more effectively.

The improved performance comes from BART’s Bayesian framework, where predictions are averaged over many trees using MCMC samples. This helps reduce variance and prevents overfitting, unlike boosting which depends heavily on tuning parameters like shrinkage.

Overall, BART achieves the best generalization among the models considered, making it the most accurate model for this dataset.

# 12. Variable Usage Count in BART

```{r chunk20}
ord <- order(bartfit$varcount.mean, decreasing = T)
bartfit$varcount.mean[ord]
```

The `varcount.mean` values show how often each predictor is used for splitting across all 1000 MCMC samples. A higher value indicates that the variable is consistently important in the model.

From the results, **LeagueN (12.755)** is the most frequently used variable, followed by **Years (12.669)**, **LeagueA (12.209)**, **CHmRun (11.959)**, and **Assists (11.934)**. Other important variables include **DivisionW, NewLeagueN, RBI, and Hits**, all of which have relatively high usage counts.

Performance-related variables such as **Walks, HmRun, CRuns, CHits, Runs, and CAtBat** also appear frequently, indicating that both career and current performance contribute to the model. Variables like **PutOuts, CRBI, CWalks, and AtBat** have slightly lower usage but still play a role.

Overall, the model relies on a mix of league indicators, experience (Years), and performance statistics. This suggests that salary is influenced not only by player performance but also by league-related factors, and no single variable dominates across all trees.

---
```{r chunk21}
cat("Boosting Test MSE (lambda = 0.001):", round(mse_boost_default, 6), "\n")

cat("Boosting Test MSE (lambda = 0.2):", round(mse_boost_large, 6), "\n")

mse_bart <- mean((ytest - yhat.bart)^2)
cat("BART Test MSE:", round(mse_bart, 6), "\n")
```

# Summary and Comparison

In this lab, we applied gradient boosting and BART to the Hitters regression problem and boosting to the Heart classification problem using the given datasets.

For the Hitters data, boosting with the default learning rate (λ = 0.001) gave a test MSE of **0.2838**. Increasing the learning rate to λ = 0.2 actually increased the test MSE to **0.3044**, showing worse performance. This suggests that a smaller learning rate works better here, as it helps the model learn more gradually and generalize better.

The relative influence results show that **CRBI** is the most important predictor, followed by variables like CHits, CHmRun, and Assists. The partial dependence plots also indicate that career performance variables have a strong impact on salary, although the relationships are nonlinear and tend to level off at higher values.

For the Heart data, both depth = 1 and depth = 2 models gave the same test error (**0.1745**). This suggests that simple trees are already sufficient, and adding interactions does not improve performance.

BART achieved the lowest test MSE (**0.2270**), outperforming boosting. This shows that BART provides better generalization, likely due to its Bayesian approach and averaging over many trees. The variable usage counts also show that multiple predictors are important, including **LeagueN, Years, CHmRun, Assists, and Hits**, indicating that both league-related factors and performance measures contribute to salary prediction.

Overall, BART performed best for regression, while simpler boosting models were sufficient for classification.
